---
title: "Capstone Group 2"
author: ""
date: "6/27/2022"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")

library(tidyverse)
library(survival)
library(survminer)
library(Hmisc)
library(forcats)
library(gtsummary)
```

# Load data

```{r}
w1 <- read.csv("wave1.csv")
w2 <- read.csv("wave2.csv")
w1$wave <- "wave1"
w2$wave <- "wave2"
wave <- rbind(w1, w2)
```

# Data cleaning

```{r}
# chronic liver disease or gastrointestinal conditions
# - cirrhosis
# - hepatitis
# - ibd(inflammatory bowel disease)

wave_clean <- wave %>%
  # cirrhosis
  mutate(cirrhosis_factor = ifelse(cirrhosis %in% 'Yes', 
                                   1, 
                                   ifelse(cirrhosis == '', NA, 0))) %>%
  # inflammatory bowel disease
  mutate(ibd_factor = ifelse(ibd %in% 'Yes', 
                             1, 
                             ifelse(ibd == '', NA, 0))) %>%
  # hepatitis
  mutate(hep_factor = ifelse(apply(sapply(c('Hepatitis C', 'Hepatitis B'), 
                                          grepl, 
                                          hepatitis), 
                                   MARGIN =  1, 
                                   FUN = any), 
                             1, 
                             ifelse(hepatitis == '', NA, 0))) %>%
  mutate(gi_condition = ifelse(cirrhosis_factor == 1 | 
                               ibd_factor == 1 | 
                               hep_factor == 1, 1, 0))

# relevel & label
wave_clean$cirrhosis_factor <- factor(wave_clean$cirrhosis_factor, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$ibd_factor <- factor(wave_clean$ibd_factor, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$hep_factor <- factor(wave_clean$hep_factor, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$gi_condition <- factor(wave_clean$gi_condition, levels=c(0, 1), labels=c("No", "Yes"))
label(wave_clean$cirrhosis_factor) <- "Cirrhosis"
label(wave_clean$ibd_factor) <- "Inflammatory Bowel Disease"
label(wave_clean$hep_factor) <- "Hepatitis"
label(wave_clean$gi_condition) <- "Chronic GI Condition"
```

```{r}
# COVID GI Symptoms and In-hospital life threatening complications
# - symptoms
# - complications

wave_clean <- wave_clean %>%
         # GI symptoms
  mutate(gi_symptoms = ifelse(grepl("Abdominal Pain|Nausea|Diarrhea", 
                                    symptoms), 
                              1, 
                              ifelse(symptoms == "", NA, 0)),
         # In-hospital life threatening complications
         life_threat_comp = ifelse(complications != "None" & 
                                   complications != "",
                                   1, 
                                   ifelse(complications == "", NA, 0))) %>%
  mutate(comp_factor = ifelse(life_threat_comp == 1 | 
                              !(vte %in% c("No", "No | No", "")) | 
                              renal_replacement == "Yes", 1, 
                              ifelse(is.na(life_threat_comp) &
                                       vte == "" &
                                       renal_replacement == "", NA, 0)))

wave_clean <- wave_clean %>%
  mutate(septic_shock = if_else(grepl("Septic Shock", complications), 
                                1, 
                                ifelse(complications == '', NA, 0)), 
         arrhythmias = if_else(grepl("Arrythmia", complications), 
                                1, 
                                ifelse(complications == '', NA, 0)),
         mi = ifelse(grepl('Myocardial Infarction', complications), 
                                1, 
                                ifelse(complications == '', NA, 0)), 
         hf1 = ifelse(grepl("Heart failure", complications), 
                                1, 
                                ifelse(complications == '', NA, 0)), 
         vte1 = ifelse(vte %in% c("No", "No | No"), 0,
                       ifelse(vte == "", NA, 1)), 
         renal_replacement1 = if_else(renal_replacement == "Yes", 
                                     1, 
                                     ifelse(renal_replacement == "", NA, 0))) 


wave_clean$gi_symptoms <- factor(wave_clean$gi_symptoms, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$comp_factor <- factor(wave_clean$comp_factor, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$septic_shock <- factor(wave_clean$septic_shock, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$arrhythmias <- factor(wave_clean$arrhythmias, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$mi <- factor(wave_clean$mi, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$hf1 <- factor(wave_clean$hf1, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$vte1 <- factor(wave_clean$vte1, levels=c(0, 1), labels=c("No", "Yes"))
wave_clean$renal_replacement1 <- factor(wave_clean$renal_replacement1, levels=c(0, 1), labels=c("No", "Yes"))
label(wave_clean$gi_symptoms) <- "Covid GI symptoms"
label(wave_clean$comp_factor) <- "Life Threatening Complications"
label(wave_clean$septic_shock) <- "Septic Shock"
label(wave_clean$arrhythmias) <- "Arrhythmias"
label(wave_clean$mi) <- "Myocardial Infarction"
label(wave_clean$hf1) <- "Heart Failure"
label(wave_clean$vte1) <- "Venous Thrombus Embolism"
label(wave_clean$renal_replacement1) <- "Renal Replacement"
```

```{r}
# Covariates
# - Comorbidities:
#   - htn
#   - cad
#   - hf
#   - cva
#   - dm
# - active_cancer
# - BMI (obesity)
# - sex
# - race_eth
# - age

wave_clean <- wave_clean %>%
  # cancer
  mutate(cancer_factor = ifelse(apply(sapply(c('Liquid', 'Solid'), 
                                             grepl, 
                                             active_cancer), 
                                      MARGIN =  1, 
                                      FUN = any), 
                                1, 
                                ifelse(active_cancer == '', NA, 0))) %>%
  # obesity: BMI >= 30
  mutate(bmi = ifelse(bmi == '33.8 | 33.8', 33.8, bmi), 
         bmi = ifelse(bmi == '24.4 | 24.4', 24.4, bmi)) %>%
  mutate(obesity = ifelse(bmi >= 30, 1, ifelse(bmi == '', NA, 0))) %>%
  # comorbidities
  mutate(dm_factor = ifelse(grepl("Yes", dm), 
                            1, 
                            ifelse(dm == '', NA, 0))) %>%
  mutate(htn_factor = ifelse(grepl("Yes", htn),
                             1,
                             ifelse(htn == "", NA, 0))) %>%
  mutate(cad_factor = ifelse(grepl("Yes", cad),
                             1,
                             ifelse(cad == "", NA, 0))) %>%
  mutate(hf_factor = ifelse(grepl("No", hf),
                            0,
                            ifelse(hf == "", NA, 1))) %>%
  mutate(cva_factor = ifelse(cva %in% "Yes",
                             1,
                             ifelse(cva == "", NA, 0))) %>%
  mutate(cva_comorbidity = ifelse(dm_factor == 1 | 
                                  htn_factor == 1 | 
                                  cad_factor == 1 | 
                                  hf_factor == 1 | 
                                  cva_factor == 1,
                                  1, 0)) %>%
  # sex
  mutate(sex = ifelse(sex == 'Female | Female', 
                      'Female',
                      ifelse(sex == '', NA, sex)))

# relevel & label

wave_clean$sex <- factor(wave_clean$sex, levels=c("Male","Female"))
label(wave_clean$sex) <- "Sex"

wave_clean$race_eth <- factor(wave_clean$race_eth, levels=c("4", "2", "3", "1", "5"))
levels(wave_clean$race_eth) <- c("White", "Black", "Asian", "Hispanic", "Other")
label(wave_clean$race_eth) <- "Race"

wave_clean$age_cat <- factor(wave_clean$age_cat, levels=c("1", "2", "3", "4", "5", "6"))
levels(wave_clean$age_cat) <- c("18 - 25", "26 - 40", "41 - 55", "56 - 70", "71 - 85", "86+")
label(wave_clean$age_cat) <- "Age"

wave_clean$cancer_factor <- factor(wave_clean$cancer_factor, levels=c(0, 1), labels=c("No", "Yes"))
label(wave_clean$cancer_factor) <- "Active Cancer"

wave_clean$cva_comorbidity <- factor(wave_clean$cva_comorbidity, levels=c(0, 1), labels=c("No", "Yes"))
label(wave_clean$cva_comorbidity) <- "Cardio Comorbidity"

wave_clean$obesity <- factor(wave_clean$obesity, levels=c(0, 1), labels=c("No", "Yes"))
label(wave_clean$obesity) <- "Obesity"
```

```{r}
# survival
wave_clean <- wave_clean %>% 
  mutate(ed_dt = as.Date(ed_dt, format = '%m/%d/%Y'),
         intubation1_dt = as.Date(intubation1_dt, format = '%m/%d/%Y'),
         death_dt = as.Date(death_dt, format = '%m/%d/%Y'),
         discharge_dt = as.Date(discharge_dt, format = '%m/%d/%Y'),
         transfer_out_dt = as.Date(transfer_out_dt, format = '%m/%d/%Y'),
         discharge_transfer_dt = if_else(is.na(discharge_dt), 
                                         transfer_out_dt, 
                                         discharge_dt),
         censor_death_dt = if_else(grepl("Yes", death), 
                                   death_dt, 
                                   discharge_transfer_dt),
         ## time to death 
         censer_death = as.numeric(difftime(censor_death_dt, ed_dt, 
                                            unit = 'days')),
         censor_intuba_dt = if_else(grepl("Yes", intubation1),
                                    intubation1_dt, 
                                    discharge_transfer_dt),
         ## time to intubation 
         censer_intuba = as.numeric(difftime(censor_intuba_dt, ed_dt, 
                                             unit = 'days'))) %>% 
  mutate(intubation1 = ifelse(grepl("Yes", intubation1),
                              1,
                              ifelse(intubation1 == "", NA, 0)),
         death = ifelse(grepl("Yes", death),
                        1,
                        ifelse(death == "", NA, 0)))

wave_clean$death <- factor(wave_clean$death, levels=c(0, 1), labels=c("No", "Yes"))
label(wave_clean$death) <- "Death"
label(wave_clean$censer_death) <- "Time to Death"
wave_clean$intubation1 <- factor(wave_clean$intubation1, levels=c(0, 1), labels=c("No", "Yes"))
label(wave_clean$intubation1) <- "Intubation"
label(wave_clean$censer_intuba) <- "Time to intubation"
```

# Table 1

```{r}
# Descriptive stats by gi_condition
obj1_stats <- wave_clean %>%
  select(sex, race_eth, age_cat, obesity, cancer_factor, cva_comorbidity,  gi_condition, gi_symptoms, comp_factor, intubation1, death, wave, septic_shock, arrhythmias, mi, hf1, renal_replacement1, vte1) %>%
  mutate(gi_condition = factor(gi_condition) %>%
           fct_explicit_na(na_level = "Missing")) %>%
  tbl_summary(by = gi_condition,  missing = 'ifany',
              label = list(age_cat ~ "Age",
                           race_eth ~ "Race")) %>%
  add_overall() %>%
  modify_caption("**Table 1. By GI Condition**") %>%
  modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Chronic GI conditions**")

# Compute p-value
obj1_p <- wave_clean %>%
  select(sex, race_eth, age_cat, obesity, cancer_factor, cva_comorbidity,  gi_condition, gi_symptoms, comp_factor, intubation1, death, wave, septic_shock, arrhythmias, mi, hf1, renal_replacement1, vte1) %>%
  tbl_summary(by = gi_condition,  missing = 'ifany',
              label = list(age_cat ~ "Age",
                           race_eth ~ "Race")) %>%
  add_p(age_cat ~ "chisq.test") %>%
  modify_column_hide(all_stat_cols())

# Create table 1
obj1_tb1 <- tbl_merge(list(obj1_stats, obj1_p)) %>%
  modify_spanning_header(everything() ~ NA)
obj1_tb1
```

```{r}
# descriptive stats by gi_symptoms
obj2_stats <- wave_clean %>%
  select(sex, race_eth, age_cat, obesity, cancer_factor, cva_comorbidity, gi_symptoms, comp_factor, intubation1, death, wave) %>%
  mutate(gi_symptoms = factor(gi_symptoms) %>%
           fct_explicit_na(na_level = "Missing")) %>%
  tbl_summary(by = gi_symptoms, missing = 'ifany',
              label = list(age_cat ~ "Age",
                           race_eth ~ "Race",
                           intubation1 ~ "Intubation")) %>%
  add_overall() %>%
  modify_caption("**Table 2. By GI Symptoms**") %>%
  modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**GI symptoms**")

# Compute p-value
obj2_p <- wave_clean %>%
  select(sex, race_eth, age_cat, obesity, cancer_factor, cva_comorbidity, gi_symptoms, comp_factor, intubation1, death, wave) %>%
  tbl_summary(by = gi_symptoms, missing = 'ifany',
              label = list(age_cat ~ "Age",
                           race_eth ~ "Race",
                           intubation1 ~ "Intubation")) %>%
  add_p(age_cat ~ "chisq.test") %>%
  modify_column_hide(all_stat_cols())

# Create table 1
obj2_tb1 <- tbl_merge(list(obj2_stats, obj2_p)) %>%
  modify_spanning_header(everything() ~ NA)
obj2_tb1
```

```{r}
# descriptive stats by wave
overall_tb1 <- wave_clean %>%
  select(sex, race_eth, age_cat, obesity, cancer_factor, cva_comorbidity, gi_condition, gi_symptoms, comp_factor, intubation1, death, wave) %>%
  tbl_summary(by = wave,  missing = 'ifany',
              label = list(age_cat ~ "Age",
                           race_eth ~ "Race")) %>%
  add_p() %>%
  add_overall() %>%
  modify_caption("**Table 1. Overall**") 
overall_tb1
```

# Objective 1

**Determine if patients who were hospitalized with GI/Hepatic conditions during the first two waves of COVID 19 were more likely to report having GI symptoms.**

Primary exposure: Presence of chronic liver disease or chronic gastrointestinal conditions, including hepatitis B, hepatitis C, cirrhosis, liver transplants, or inflammatory bowel disease. 

Primary outcome: COVID Symptoms, including abdominal pain, nausea or diarrhea.

## Logistic model
```{r}
fit_obj1 <- glm(gi_symptoms ~ gi_condition + sex + race_eth + age_cat + obesity + cancer_factor + cva_comorbidity + sex*gi_condition + wave, data = wave_clean, family = "binomial")
summary(fit_obj1)
```

```{r}
# Odds Ratio
round(exp(cbind(OR = coef(fit_obj1), confint(fit_obj1))), 3)
```

**Interpret**: The result shows that there is no significant association between GI condition and COVID GI symptoms at 5% level of significance. The odds ratio for GI condition is positive (1.178), but the confidence interval include 1. We also include the interaction term for sex because female are more likely to have GI symptoms compared to male, but the coefficient of the interaction term is not significant either.

Moreover, we found that patients who have obesity are more likely to have GI symptoms compared to patients who do not have obesity. Studies have shown that obesity is a risk factor of GI disease.

We also surprisingly found that elder people have a less possibility to have GI symptoms compared to youngsters. This could make sense because GI symptoms are milder than lung or other symptoms. Elder people might tend to have more severe symptoms instead of GI symptoms.

# Objective 2

**Determine if patients hospitalized with GI symptoms during the first two waves of COVID 19 had similar in-hospital outcomes to those who did not report such symptoms.**

Primary exposure: COVID Symptoms, including abdominal pain, nausea or diarrhea.

Primary outcome: in-hospital life threatening complications (Septic shock, positive blood culture, renal replacement therapy, arrhythmias, myocardial infarction, heart failure, venous thrombus embolism, coagulation), intubation, or death. 

## Logistic model
```{r}
wave_clean <- wave_clean %>%
  mutate(outcome = ifelse(comp_factor == "Yes" | death == "Yes" | intubation1 == "Yes", 1, 0))


fit_obj2 <- glm(outcome ~ gi_symptoms + sex + race_eth + age_cat + obesity + cancer_factor + cva_comorbidity + wave + sex * wave, data = wave_clean, family = "binomial")
summary(fit_obj2)
```

```{r}
# Odds Ratio
round(exp(cbind(OR = coef(fit_obj2), confint(fit_obj2))), 3)
```

**Interpret**: A multinational logistic regression model was fitted to investigate the relationship between development of GI symptoms and in-hospital outcomes during the first two waves of COVID 19. In-hospital outcomes are defined as in-hospital life threatening complications (Septic shock, positive blood culture, renal replacement therapy, arrhythmias, myocardial infarction, heart failure, venous thrombus embolism, coagulation), intubation, or death.

According to the model, the log of the odds of in-hospital outcomes was negatively related to GI symptoms (p < .05) and gender (p < .05). Patients with GI symptoms were more likely to have in-hospital life threatening complications, intubation, or death after controlling for other factors. Patients with GI symptoms were 0.861 times more likely to have in-hospital life-threatening complications, intubation, or death than patients without GI symptoms.

Adjusting for other variables, elder patient, patients with obesity, cancer and cardiovascular comorbidities were likely to experience the above in-hospital outcomes (p < .05).


## Death Survival
```{r}
survdiff(Surv(censer_death, as.numeric(death)) ~ gi_symptoms, data = wave_clean)

ggsurvplot(
  fit = survfit(Surv(censer_death, as.numeric(death)) ~ gi_symptoms, data = wave_clean),
  xlab = "Days",
  ylab = "Survival probability",
  pval = TRUE,
  surv.median.line = "hv",
  risk.table = TRUE
)

survfit(Surv(censer_death, as.numeric(death)) ~ gi_symptoms, data = wave_clean)
```

**Interpret**: The effect of GI symptoms on survival time was investigated using survival analysis. Patients with GI symptoms had a median survival time of 51 days, whereas those without GI symptoms had a median survival time of 35 days. There was a significant difference between the two survival plots (p < 0.001). One possible explanation for this outcome is that GI symptoms may be minor in COVID-19 patients. Some of the more significant symptoms, such as lung symptoms, may be linked to death.

# Objective 3

**Determine in-hospital complications, including mortality among those with GI/Hepatic comorbidities of Covid 19 hospitalizations in a large hospital in NYC **

Primary exposure: Presence of chronic liver disease or chronic gastrointestinal conditions, including hepatitis B, hepatitis C, cirrhosis, liver transplants, or inflammatory bowel disease

Primary outcome: in-hospital life threatening complications (Septic shock, positive blood culture, renal replacement therapy, arrhythmias, myocardial infarction, heart failure, venous thrombus embolism, coagulation), intubation, or death. 

```{r}
# descriptive stats
# obj3_tb1 <- wave_clean %>%
#   select(sex, race_eth, age_cat, obesity, cancer_factor, cva_comorbidity, gi_condition, comp_factor, intubation1, death) %>%
#   tbl_summary(by = gi_condition, missing = 'no',
#               label = list(age_cat ~ "Age",
#                            race_eth ~ "Race",
#                            intubation1 ~ "Intubation")) %>%
#   add_p() %>%
#   add_overall() %>%
#   modify_caption("**Table 3. Objective 3**") %>%
#   modify_spanning_header(c("stat_1", "stat_2") ~ "**GI condition**")
# obj3_tb1
```

## Life threatening complications

```{r}
wave_clean <- wave_clean %>%
  mutate(outcome1 = ifelse(comp_factor == "Yes" | death == "Yes", 1, 0))

```


## Logistic model
```{r}
## outcome
fit_obj3_outcome <- glm(outcome1 ~ gi_condition + sex + race_eth + age_cat + obesity + cancer_factor + cva_comorbidity + sex*gi_condition + wave, wave_clean, family = binomial)
summary(fit_obj3_outcome)

## death
fit_obj3_death <- glm(death ~ gi_condition + sex + race_eth + age_cat + obesity + cancer_factor + cva_comorbidity + wave, wave_clean, family = binomial)
summary(fit_obj3_death)

## intubation
fit_obj3_intuba <- glm(intubation1 ~ gi_condition + sex + race_eth + age_cat + obesity + cancer_factor + cva_comorbidity + wave, wave_clean, family = binomial)
summary(fit_obj3_intuba)

## complication
fit_obj3_comp <- glm(comp_factor ~ gi_condition + sex + race_eth + age_cat + obesity + cancer_factor + cva_comorbidity + wave, wave_clean, family = binomial)
summary(fit_obj3_comp)
```

```{r}
# Odds Ratio
round(exp(cbind(OR = coef(fit_obj3_outcome), confint(fit_obj3_outcome))), 3)
```

**Interpret**: GI condition is not significantly associated with in hospital outcomes, including life threatening complications, death and intubation. Looking at the hospital outcomes separately, the predictor GI condition is still not significant. 

However, we found that the results were similar in outcome death and life threatening complications for their close p values and both positive estimated coefficients. The results of intubation show a much larger p-value and a negative estimated coefficients which indicates an opposite direction. 

## Death Survival
```{r}
## death-survival
wave_clean$death_code = as.numeric(wave_clean$death)
ggsurvplot( 
  fit = survfit(Surv(censer_death,death_code) ~ gi_condition, data = wave_clean),
  xlab = "Days", 
  ylab = "Survival probability",
  pval = TRUE,
  surv.median.line = "hv",
  risk.table = TRUE)

survfit(Surv(censer_death, death_code) ~ gi_condition, data = wave_clean)
```

**Interpret**: The effect of chronic GI condition on survival time was investigated using survival analysis. Patients with chronic GI condition had a median survival time of 31 days, whereas those without chronic GI condition had a median survival time of 40 days. However, the difference was not significant between the two survival plots (p=0.72). 

# Objective 4

**Describe patient-reported outcomes (PROs) and clinical outcomes among patients with a subset of patients with GI/Hepatic conditions approximately one year after COVID admission (details are in the original proposal). **

Objective 4: Original proposal.

Primary exposure: Presence of chronic liver disease or chronic gastrointestinal conditions, including hepatitis B, hepatitis C, cirrhosis, liver transplants, or inflammatory bowel disease

Primary outcome: Patient Reported Outcomes (PROs): Post Acute Covid Sequalae such as brain fog, cough, shortness of breath, neuropathy, headache, muscle aches, diarrhea, chest pain, abdominal pain, loss of smell taste, fatigue or other.
Clinical outcomes: post hospitalization readmission or death.
