---
title: "capstone"
author: "JIN ZHU"
date: "5/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(survival)
library(survminer)
```

```{r}
w1 <- read.csv("wave1.csv")
w2 <- read.csv("wave2.csv")


wave <- rbind(w1, w2)
wave


str(w2)
wave %>% 
  mutate(ed_dt = as.Date(ed_dt, format='%m/%d/%Y'),
         intubation1_dt = as.Date(intubation1_dt, format='%m/%d/%Y'),
         death_dt = as.Date( death_dt,format='%m/%d/%Y'),
         discharge_dt = as.Date(discharge_dt,format='%m/%d/%Y'),
         transfer_out_dt = as.Date(transfer_out_dt , format = '%m/%d/%Y'))
  
  


## ards: Acute respiratory distress syndrome 
## cad: coronary artery disease
## cirrhosis 肝硬化
## rheum. Rheumatoid arthritis is a chronic inflammatory disorder
```

```{r}
# missing values?   find the exposures? exposure pivot_longer/  death: time to death  survival data 
```

# Objective 1
1.	Determine if patients who were hospitalized with GI/Hepatic conditions during the first two waves of COVID 19 were more likely to report having GI symptoms.

Primary exposure: Presence of chronic liver disease or chronic gastrointestinal conditions, including hepatitis B, hepatitis C, cirrhosis, liver transplants, or inflammatory bowel disease. 

Primary outcome: COVID Symptoms, including abdominal pain, nausea or diarrhea.



# Objective 2
2.	Determine if patients hospitalized with GI symptoms during the first two waves of COVID 19 had similar in-hospital outcomes to those who did not report such symptoms.


Primary exposure: COVID Symptoms, including abdominal pain, nausea or diarrhea.

Primary outcome: in-hospital life threatening complications (Septic shock, positive blood culture, renal replacement therapy, arrhythmias, myocardial infarction, heart failure, venous thrombus embolism, coagulation), intubation, or death. 



# Objective 3
3.	Determine in-hospital complications, including mortality among those with GI/Hepatic comorbidities of Covid 19 hospitalizations in a large hospital in NYC 

Primary exposure: Presence of chronic liver disease or chronic gastrointestinal conditions, including hepatitis B, hepatitis C, cirrhosis, liver transplants, or inflammatory bowel disease

Primary outcome: in-hospital life threatening complications (Septic shock, positive blood culture, renal replacement therapy, arrhythmias, myocardial infarction, heart failure, venous thrombus embolism, coagulation), intubation, or death. 



# Objective 4
4.	Describe patient-reported outcomes (PROs) and clinical outcomes among patients with a subset of patients with GI/Hepatic conditions approximately one year after COVID admission (details are in the original proposal). 

Objective 4: Original proposal.

Primary exposure: Presence of chronic liver disease or chronic gastrointestinal conditions, including hepatitis B, hepatitis C, cirrhosis, liver transplants, or inflammatory bowel disease

Primary outcome: Patient Reported Outcomes (PROs): Post Acute Covid Sequalae such as brain fog, cough, shortness of breath, neuropathy, headache, muscle aches, diarrhea, chest pain, abdominal pain, loss of smell taste, fatigue or other.
Clinical outcomes: post hospitalization readmission or death.




