---
title: "capstone"
author: "JIN ZHU"
date: "5/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(survival)
library(survminer)
library(Hmisc)
library(gtsummary)
```

# Load data

```{r}
w1 <- read.csv("wave1.csv")
w2 <- read.csv("wave2.csv")


wave <- rbind(w1, w2)
```

# Data cleaning

```{r}
# chronic liver disease or gastrointestinal conditions
# - cirrhosis
# - hepatitis
# - ibd(inflammatory bowel disease)

wave <- wave %>%
  # cirrhosis
  mutate(cirrhosis_factor = ifelse(cirrhosis %in% 'Yes', 
                                   1, 
                                   ifelse(cirrhosis == '', NA, 0))) %>%
  # inflammatory bowel disease
  mutate(ibd_factor = ifelse(ibd %in% 'Yes', 
                             1, 
                             ifelse(ibd == '', NA, 0))) %>%
  # hepatitis
  mutate(hep_factor = ifelse(apply(sapply(c('Hepatitis C', 'Hepatitis B'), 
                                          grepl, 
                                          hepatitis), 
                                   MARGIN =  1, 
                                   FUN = any), 
                             1, 
                             ifelse(hepatitis == '', NA, 0))) %>%
  mutate(gi_condition = ifelse(cirrhosis_factor == 1 | 
                               ibd_factor == 1 | 
                               hep_factor == 1, 1, 0))

# relevel & label

wave$cirrhosis_factor <- factor(wave$cirrhosis_factor, levels=c(0, 1), labels=c("No", "Yes"))
wave$ibd_factor <- factor(wave$ibd_factor, levels=c(0, 1), labels=c("No", "Yes"))
wave$hep_factor <- factor(wave$hep_factor, levels=c(0, 1), labels=c("No", "Yes"))
wave$gi_condition <- factor(wave$gi_condition, levels=c(0, 1), labels=c("No", "Yes"))
label(wave$cirrhosis_factor) <- "Cirrhosis"
label(wave$ibd_factor) <- "Inflammatory bowel disease"
label(wave$hep_factor) <- "Hepatitis"
label(wave$gi_condition) <- "Chronic GI Condition"
```

```{r}
# COVID GI Symptoms and In-hospital life threatening complications
# - symptoms
# - complications

wave <- wave %>%
         # GI symptoms
  mutate(gi_symptoms = ifelse(grepl("Abdominal Pain|Nausea|Diarrhea", 
                                    symptoms), 
                              1, 
                              ifelse(symptoms == "", NA, 0)),
         # In-hospital life threatening complications
         life_threat_comp = ifelse(complications != "None" & 
                                   complications != "",
                                   1, 
                                   ifelse(complications == "", NA, 0))) %>%
  mutate(comp_factor = ifelse(life_threat_comp == 1 | 
                              vte == 1 | 
                              renal_replacement == 1, 1, 0))


wave$gi_symptoms <- factor(wave$gi_symptoms, levels=c(0, 1), labels=c("No", "Yes"))
wave$comp_factor <- factor(wave$comp_factor, levels=c(0, 1), labels=c("No", "Yes"))
label(wave$gi_symptoms) <- "Covid GI symptoms"
label(wave$comp_factor) <- "Life threatening complications"
```

```{r}
# Covariates
# - Comorbidities:
#   - htn
#   - cad
#   - hf
#   - cva
#   - dm
# - active_cancer
# - BMI (obesity)
# - sex
# - race_eth
# - age

wave <- wave %>%
  # cancer
  mutate(cancer_factor = ifelse(apply(sapply(c('Liquid', 'Solid'), 
                                             grepl, 
                                             active_cancer), 
                                      MARGIN =  1, 
                                      FUN = any), 
                                1, 
                                ifelse(active_cancer == '', NA, 0))) %>%
  # obesity: BMI >= 30
  mutate(bmi = ifelse(bmi == '33.8 | 33.8', 33.8, bmi), 
         bmi = ifelse(bmi == '24.4 | 24.4', 24.4, bmi)) %>%
  mutate(obesity = ifelse(bmi >= 30, 1, ifelse(bmi == '', NA, 0))) %>%
  # comorbidities
  mutate(dm_factor = ifelse(dm %in% 'Yes', 
                            1, 
                            ifelse(dm == '', NA, 0))) %>%
  mutate(htn_factor = ifelse(htn %in% "Yes",
                             1,
                             ifelse(htn == "", NA, 0))) %>%
  mutate(cad_factor = ifelse(cad %in% "Yes",
                             1,
                             ifelse(cad == "", NA, 0))) %>%
  mutate(hf_factor = ifelse(hf %in% "No",
                            0,
                            ifelse(hf == "", NA, 1))) %>%
  mutate(cva_factor = ifelse(cva %in% "Yes",
                             1,
                             ifelse(cva == "", NA, 0))) %>%
  mutate(comorbidity = ifelse(dm_factor == 1 | 
                              htn_factor == 1 | 
                              cad_factor == 1 | 
                              hf_factor == 1 | 
                              cva_factor == 1,
                              1, 0)) %>%
  # sex
  mutate(sex = ifelse(sex == 'Female | Female', 
                      'Female',
                      ifelse(sex == '', NA, sex)))

# relevel & label

wave$sex <- factor(wave$sex, levels=c("Male","Female"))
label(wave$sex) <- "Sex"

wave$race_eth <- factor(wave$race_eth, levels=c("1", "2", "3", "4", "5"))
levels(wave$race_eth) <- c("Hispanic", "Black", "Asian", "White", "Other")
label(wave$race_eth) <- "Race"

wave$age_cat <-  factor(wave$age_cat, levels=c("1", "2", "3", "4", "5", "6"))
levels(wave$age_cat) <- c("18 – 25", "26 – 40", "41 - 55", "56 – 70", "70 - 85", "86+")
label(wave$age_cat) <- "Age"

wave$cancer_factor <- factor(wave$cancer_factor, levels=c(0, 1), labels=c("No", "Yes"))
label(wave$cancer_factor) <- "Active Cancer"

wave$comorbidity <- factor(wave$comorbidity, levels=c(0, 1), labels=c("No", "Yes"))
label(wave$comorbidity) <- "Comorbidity"

wave$obesity <- factor(wave$obesity, levels=c(0, 1), labels=c("No", "Yes"))
label(wave$obesity) <- "Obesity"
```

```{r}
# survival
wave <- wave %>% 
  mutate(ed_dt = as.Date(ed_dt, format = '%m/%d/%Y'),
         intubation1_dt = as.Date(intubation1_dt, format = '%m/%d/%Y'),
         death_dt = as.Date(death_dt, format = '%m/%d/%Y'),
         discharge_dt = as.Date(discharge_dt, format = '%m/%d/%Y'),
         transfer_out_dt = as.Date(transfer_out_dt, format = '%m/%d/%Y'),
         discharge_transfer_dt = if_else(is.na(discharge_dt), 
                                         transfer_out_dt, 
                                         discharge_dt),
         censor_death_dt = if_else(death == 'Yes', 
                                   death_dt, 
                                   discharge_transfer_dt),
         ## time to death 
         censer_death = as.numeric(difftime(censor_death_dt, ed_dt, 
                                            unit = 'days')),
         censor_intuba_dt = if_else(intubation1 =='Yes',
                                    intubation1_dt, 
                                    discharge_transfer_dt),
         ## time to intubation 
         censer_intuba = as.numeric(difftime(censor_intuba_dt, ed_dt, 
                                             unit = 'days'))) %>% 
  mutate(intubation1 = ifelse(intubation1 %in% "Yes",
                              1,
                              ifelse(intubation1 == "", NA, 0)),
         death = ifelse(death %in% "Yes",
                        1,
                        ifelse(death == "", NA, 0)))

wave$death <- factor(wave$death, levels=c(0, 1), labels=c("No", "Yes"))
label(wave$death) <- "Death"
wave$intubation1 <- factor(wave$intubation1, levels=c(0, 1), labels=c("No", "Yes"))
label(wave$intubation1) <- "Intubation"
```


```{r}
# ### discharge & transfer out 
# wave %>%
#   select(intubation1,intubation1_dt,discharge_dt, transfer_out_dt) %>%
#   filter(discharge_dt !=''& transfer_out_dt == '')
# ###
# 
# ## 932 patients with no discharge date and no transfer_out_dt; 90 patients tranfer out; no both records for the two variables 
# 
# wave %>%
#   select(complications,  death, intubation1 ) ## not mutullay exclusive, seperate outcomes 
# 
# wave %>%
#   select(intubation1, discharge_dt,intubation1_dt )%>% 
#   filter(intubation1 == 'Yes' & intubation1_dt == '')
# 
# wave %>%
#   select( death,  death_dt,discharge_dt ) 
# 
# wave %>% ## survival data used 
#   select(intubation1,  censer_intuba, death, censer_death)
```

```{r}
# missing values?   find the exposures? exposure pivot_longer/  death: time to death  survival data 
```

# Cleaned dataset 

```{r}
wave_clean <- wave

# Death
# wave_clean %>% 
#   filter(!is.na(death))
#
# Intubation
# wave_clean %>%
#   filter(!is.na(intubation1))
```

# Objective 1

1.	Determine if patients who were hospitalized with GI/Hepatic conditions during the first two waves of COVID 19 were more likely to report having GI symptoms.

Primary exposure: Presence of chronic liver disease or chronic gastrointestinal conditions, including hepatitis B, hepatitis C, cirrhosis, liver transplants, or inflammatory bowel disease. 
zz
Primary outcome: COVID Symptoms, including abdominal pain, nausea or diarrhea.

```{r}
# descriptive stats
obj1_tb1 <- wave_clean %>%
  select(sex, race_eth, age_cat, obesity, cancer_factor, comorbidity, gi_condition, gi_symptoms) %>%
  tbl_summary(by = gi_condition,  missing = 'no', 
              label = list(age_cat ~ "Age", 
                           race_eth ~ "Race")) %>%
  add_p() %>%
  add_overall() %>%
  modify_caption("**Table 1. Objective 1**") %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Chronic GI conditions**")
obj1_tb1
```

# Objective 2

2.	Determine if patients hospitalized with GI symptoms during the first two waves of COVID 19 had similar in-hospital outcomes to those who did not report such symptoms.

Primary exposure: COVID Symptoms, including abdominal pain, nausea or diarrhea.

Primary outcome: in-hospital life threatening complications (Septic shock, positive blood culture, renal replacement therapy, arrhythmias, myocardial infarction, heart failure, venous thrombus embolism, coagulation), intubation, or death. 

```{r}
# descriptive stats
obj2_tb1 <- wave_clean %>%
  select(sex, race_eth, age_cat, obesity, cancer_factor, comorbidity, gi_symptoms, comp_factor, intubation1, death) %>%
  tbl_summary(by = gi_symptoms, missing = 'no', 
              label = list(age_cat ~ "Age", 
                           race_eth ~ "Race",
                           intubation1 ~ "Intubation")) %>%
  add_p() %>%
  add_overall() %>%
  modify_caption("**Table 2. Objective 2**") %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**GI symptoms**")
obj2_tb1
```


```{r}
# wave %>% ## survival data used 
#   select(gi_symptoms, intubation1,  censer_intuba, death, censer_death)


survdiff(Surv(censer_intuba, intubation1) ~ gi_symptoms, data = wave_clean)

## intubation 
ggsurvplot( 
  fit = survfit(Surv(censer_intuba, intubation1) ~ gi_symptoms, data = wave_clean),
  xlab = "Days", 
  ylab = "No intubation probability",
  pval = TRUE,
  surv.median.line = "hv",
  risk.table = TRUE)

## Survival 
ggsurvplot( 
  fit = survfit(Surv(censer_death, death) ~ gi_symptoms, data = wave_clean),
  xlab = "Days", 
  ylab = "Survival probability",
  pval = TRUE,
  surv.median.line = "hv",
  risk.table = TRUE)

```

# Objective 3

3.	Determine in-hospital complications, including mortality among those with GI/Hepatic comorbidities of Covid 19 hospitalizations in a large hospital in NYC 

Primary exposure: Presence of chronic liver disease or chronic gastrointestinal conditions, including hepatitis B, hepatitis C, cirrhosis, liver transplants, or inflammatory bowel disease

Primary outcome: in-hospital life threatening complications (Septic shock, positive blood culture, renal replacement therapy, arrhythmias, myocardial infarction, heart failure, venous thrombus embolism, coagulation), intubation, or death. 

```{r}
# descriptive stats
obj3_tb1 <- wave_clean %>%
  select(sex, race_eth, age_cat, obesity, cancer_factor, comorbidity, gi_condition, comp_factor, intubation1, death) %>%
  tbl_summary(by = gi_condition,  missing = 'no', 
              label = list(age_cat ~ "Age", 
                           race_eth ~ "Race",
                           intubation1 ~ "Intubation")) %>%
  add_p() %>%
  add_overall() %>%
  modify_caption("**Table 3. Objective 3**") %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**GI condition**")
obj3_tb1
```


```{r}
## death
ggsurvplot( 
  fit = survfit(Surv(censer_death, death) ~ gi_condition, data = wave_clean),
  xlab = "Days", 
  ylab = "Survival probability",
  pval = TRUE,
  surv.median.line = "hv",
  risk.table = TRUE)

## intubation 
ggsurvplot( 
  fit = survfit(Surv(censer_intuba, intubation1) ~ gi_condition, data = wave_clean),
  xlab = "Days", 
  ylab = "No intubation probability",
  pval = TRUE,
  surv.median.line = "hv",
  risk.table = TRUE)


```

# Objective 4

4.	Describe patient-reported outcomes (PROs) and clinical outcomes among patients with a subset of patients with GI/Hepatic conditions approximately one year after COVID admission (details are in the original proposal). 

Objective 4: Original proposal.

Primary exposure: Presence of chronic liver disease or chronic gastrointestinal conditions, including hepatitis B, hepatitis C, cirrhosis, liver transplants, or inflammatory bowel disease

Primary outcome: Patient Reported Outcomes (PROs): Post Acute Covid Sequalae such as brain fog, cough, shortness of breath, neuropathy, headache, muscle aches, diarrhea, chest pain, abdominal pain, loss of smell taste, fatigue or other.
Clinical outcomes: post hospitalization readmission or death.




